Early warning signals: The charted and uncharted territories

### 2013-05-17 manuscript.tex
Updated tex and pdf #26 looks like some eps images may need to be resized (issue #27) as they get embedded without resizing

\documentclass[author-year, review]{elsarticle} %review=doublespace preprint=single 5p=2 column
%\documentclass[author-year, review]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add 
%\linenumbers % turns line numbering on 
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

% A modified page layout
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Early warning signals: The charted and uncharted territories},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\begin{document}
\begin{frontmatter}
  \title{Early warning signals: The charted and uncharted territories}
  \author[cstar]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@gmail.com}
  \author[esp]{Noam Ross\corref{cor1}}
  \ead{noam.ross@gmail.com}
  \author[esp]{Alan Hastings}
  \cortext[cor1]{authors contributed equally}
  \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
  \address[esp]{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA}

@@

The realization that complex systems such as ecological communities can collapse or shift regimes suddenly and without rapid external forcing poses a serious challenge to our understanding and management of the natural world.  The potential to identify early warning signals that would allow researchers and managers to predict such events before they happen has therefore been an invaluable discovery that offers a way forward in spite of such seemingly unpredictable behavior.  Research into early warning signals has demonstrated that it is possible to define and detect such early warning signals in advance of a transition in certain contexts.  Here we describe the pattern emerging as research continues to explore just how far we can generalize these results.  A core of examples emerges that shares three properties: the phenomenon of rapid regime shifts,  a pattern of 'critical slowing down' that can be used to detect the approaching shift, and a mechanism of bifurcation driving the sudden change.  As research has expanded beyond these core examples, it is becoming clear that not all systems that show regime shifts exhibit critical slowing down, or vice versa.  Even when systems exhibit critical slowing down, statistical detection is a challenge.  We review the literature that explores these edge cases and highlight the need for (a) new early warning behaviors that can be used in cases where rapid shifts do not exhibit critical slowing down, (b) the development of methods to identify which behavior might be an appropriate signal when encountering a novel system; bearing in mind that a positive indication for some systems is a negative indication in others, and (c) statistical methods that can distinguish between signatures of early warning behaviors and noise.
  \end{abstract}
  \begin{keyword}
early warning signals \sep regime shifts \sep bifurcation \sep critical slowing down 
   \end{keyword}

 \end{frontmatter}

@@

Many natural systems exhibit regime shifts - rapid changes in the state
and conditions of system behavior. Such shifts include lake
eutrophication (Carpenter, Ludwig, and Brock 1999), algal overgrowth of
coral systems (Mumby, Hastings, and Edwards 2007), fishery collapse
(Jackson et al. 2001), desertification of grasslands (Kéfi et al. 2007),
and rapid changes in climate (Dakos et al. 2008; Lenton et al. 2009).
Such dramatic shifts have the potential to impact ecosystem health and
human well-being. Thus, it is important to develop strategies for
adaptation, mitigation, and avoidance of such shifts.

The idea that complex systems such as ecosystems could change suddenly
and without warning goes back to the 1960s (Lewontin 1969; Holling 1973;
May 1977). Such early work revealed that even simple models with the
appropriate nonlinearities were capable of unpredictable behavior. The
only way to predict the transition was to have the right model -- and
that meant having already had the chance to observe the transition (even
then, this remains a tough problem). One cogent early example (Ludwig,
Jones, and Holling 1978) demonstrated how knowledge of the forms and
time scales of interactions among insects, birds, and trees could lead
to a qualitative model that essentially predicted the possibility of
regime shifts.

@@

Foundational research in EWS identified certain patterns that may
forecast a sudden transition in a wide variety of systems (Scheffer et
al. 2009). Most extensively researched is the phenomenon of critical
slowing down (CSD), which is manifested as a pattern of increasing
variance or autocorrelation of a system. Subsequent work has begun to
identify a growing library of cases in which these indicators are not
present before a transition (Hastings and Wysham 2010; Schreiber and
Rudolf 2008; Bel, Hagberg, and Meron 2012; Schreiber and Rittenhouse
2004), or are observed in the absence of any transition (Kéfi et al.
2012). These examples are distinct from the more well-known case of
statistical error -- such as a signal that is present but too weak to
detect due to insufficient available data (see Dakos et al. (2008);
Scheffer et al. (2009) and Perretti and Munch (2012)). Instead, such
work moves into new territory where different underlying mechanisms have
lead to starkly different patterns. Determining which underlying
mechanisms are present is a substantial empirical and theoretical
challenge. When does critical slowing down correspond to the assumptions
made?

Here we review a variety of mechanisms that may lead to rapid (or
``catastrophic'') regime shifts in ecological systems, as well as
mechanisms that generate early warning signals. We focus on CSD and its
manifestations as they are the most commonly studied warning signals. We
illustrate that not all rapid shifts exhibit CSD, and not all
observations of CSD involve rapid shifts. Thus the issue of determining
EWS is really two-fold: first, identify classes of systems where the
warning signal is expected and conversely systems that may undergo
shifts without such signals, and second, to determine appropriate
statistical tools to detect the warning signal. In this paper we review
both aspects of the overall question.

\begin{longtable}[c]{@{}ll@{}}
\hline\noalign{\medskip}
\begin{minipage}[t]{0.44\columnwidth}\raggedright
\textbf{Critical slowing down (CSD)}
\end{minipage} & \begin{minipage}[t]{0.56\columnwidth}\raggedright
A system's slowing response to perturbations as it's dominant eigenvalue
approaches zero, often expressed in greater variance, autocorrelation,
and return time. CSD is one possible EWS.
\end{minipage}
\\\noalign{\medskip}
\begin{minipage}[t]{0.44\columnwidth}\raggedright
\textbf{Early warning signals (EWS)}
\end{minipage} & \begin{minipage}[t]{0.56\columnwidth}\raggedright
A general term for dynamic patterns in system behavior that precede
regime shifts. Though CSD phenomena are among the best studied EWS, some
shifts will require alternative signals; Figure 1.
\end{minipage}
\\\noalign{\medskip}
\hline
\end{longtable}

@@

CSD has been studied extensively in theoretical (Wissel 1984; Gandhi,
Levin, and Orszag 1998; Carpenter and Brock 2006; Hastings and Wysham
2010; Lade and Gross 2012; V. Dakos, Kéfi, et al. 2011; C. Boettiger and
Hastings 2012) and empirical contexts (Drake and Griffen 2010; J.
Carpenter 2011; Veraart et al. 2011; Dai et al. 2012; Wang et al. 2012)
as a potential EWS for regime shifts. CSD occurs as a system's dominant
eigenvalue approaches zero due to a changing (possibly deteriorating)
environment. As the eigenvalue approaches zero, the system's response to
small perturbations slows. This change in dynamic properties of a system
can be expressed in greater variance, autocorrelation, and return time
of observed state variables.

@@

\begin{figure}[htbp]
\centering
\includegraphics{ews-venn.pdf}
\caption{Venn diagram representing the intersecting domains of rapid
regime shifts, bifurcations, and critical slowing down. Labels in italic
are example phenomena that occur in each domain. Roman numerals indicate
example literature (right) exploring each domain, and also refer to
sections below describing those domains. The center domain (I) where all
three phenomena intersect, is the most extensively researched domain of
the EWS field. Literature outside this charted region does not yet
provide the needed EWS, but hints where existing signals based on CSD
may be insufficient or misleading.}
\end{figure}

@@

Much of the (most visible) recent research in EWS has focused on the
center of the diagram, where all three concepts intersect. The warning
signal patterns postulated, such as increasing variance and coefficient
of variation, (Carpenter and Brock 2006), increasing autocorrelation
(Dakos et al. 2008), increasing skewness (Guttal and Jayaprakash 2008)
can all be directly derived from the changing eigenvalue in a saddle
node (also called fold) bifurcation. Consequently, experimental
evaluations of warning signals have largely focused on this situation as
well. CSD has frequently been studied in the context of models
exhibiting saddle-node bifurcations.

Dai et al. (2012) studied yeast cell growth in a microcosm and
demonstrated that an Allee effect created a saddle-node bifurcation in
the system. When the cell density was reduced to levels near the
bifurcation point, a decrease in recovery time (increase in variance and
autocorrelation over time) was observed. Veraart et al. (2011) studied a
system of cyanobacteria where models suggest a saddle-node bifurcation
driven by light inhibition. They also found increases in autocorrelation
and decreased recovery rates as the system approached the bifurcation.
These important experiments are among the best demonstrations that
saddle-node bifurcation dynamics really occur in natural systems, and
can be accompanied by reliable detection of EWS, at least when
sufficient data sampling, replicates, and controls are available.

Carpenter et al. (2011) provide a larger-scale example in which a lake
ecosystem is manipulated towards a sudden transition through the
introduction of a predator, while a neighboring experimental lake
provides a control. In this and similar lake systems, bifurcation is
thought to be driven in part by trophic interactions where adult fish
prey on the competitors of their juveniles (Carpenter and Kitchell 1996;
Walters2001; Carpenter et al. 2008) which leads to a saddle-node
bifurcation. While the underlying dynamics of a whole lake ecosystem are
less tractable than the laboratory controlled chemoststats of
microorganisms, the system is understood well enough to anticipate that
a sudden transition can be induced under the intended manipulation. Like
the laboratory examples, this helps eliminate the options outside the
circle ``bifurcations,'' in Figure 1. The observed warning signals then
place it in the center of the diagram.

@@

These are some of the most problematic cases. They represent disruptive
but potentially avoidable events, but would not be detected by using CSD
as an EWS. These cases include bifurcations in continuous time
{[}@Schrieber2008{]} and discrete time {[}@Schrieber2003{]}, explicitly
spatial (Bel, Hagberg, and Meron 2012) and non-spatial, chaotic
(Schreiber 2003, ) and non-chaotic (Hastings and Wysham 2010, , )
examples. Before warning signals can be reliably applied to novel
systems, research must provide a way to discern if the dynamics
correspond to the better understood warning signals of the saddle-node
case or the more complex patterns such as the examples discussed here.

One class of bifurcations in which we would not expect to see CSD prior
to regime shift are sometimes known as \emph{crises}. Crises are sudden
changes in the dynamics of chaotic attractors that occur in response to
small changes in parameters (Grebogi, Ott, and Yorke 1983). Chaotic
attractors are features of many ecological models (Hastings et al.
1993), and chaotic behavior has been shown in some ecological systems
{[}@Constantino1997{]}.

@@

\begin{figure}[htbp]
\centering
\includegraphics{schreiber-fig.pdf}
\caption{A system where variance decreases prior to a population
collapse; adapted from Schreiber (2003). In this model, prey species
with high growth rates exhibit chaotic dynamics under predation, but
populations collapse when predation increases beyond a threshold value.
Left: The population level as a function of predation rate. Mean
dynamics shown as black line, realizations with varying initial
conditions shown as grey dots; see Schreiber (2003). Middle: Variance of
the prey population level. Note that it \emph{decreases} as predation
rate approaches the threshold. Right: Lag-1 Autocorrelation in prey
population dynamics increases as the threshold is approached}
\end{figure}

@@

Another non-chaotic example is found in some spatially extended systems
that exhibit a type of bifurcation not accompanied by CSD. In this class
of models, individual locations are subject to saddle node-type regime
shifts and influence adjacent locations via short-range facilitation and
long-range competition. Such models are used represent transitions
between vegetation types in response to changing water availability, and
reproduce naturally occurring vegetation patterns (Rietkerk and van de
Koppel 2008). In such systems, a regime shift in one location can
propagate spatially and transition the whole system from one regime to
another. Such a transition occurs if the control parameter (e.g.,
rainfall), exceeds the \emph{Maxwell point} - the value at which a local
disturbance propagates outwards (Bel, Hagberg, and Meron 2012). The
Maxwell point may be far from the level at which an individual location
would undergo a saddle-node bifurcation, and thus the system's global
dynamics would not exhibit CSD prior to such a transition. This case
illustrates the importance of distinguishing between \emph{local} and
\emph{global} system dynamics and identifying the appropriate scale of
observation.

Finally, Boerlijst, Oudman, and de Roos (2013) found that indicators of
CSD do not appear prior to saddle-node bifurcations when perturbations
are not in the direction of a system's dominant eigenvalue, and even
then may only appear in one variable of the system. In their example
case, increased variance and autocorrelation only occurred when noise
was applied to the juvenile population of a model with juveniles,
adults, and predators, and it did not appear when identical noise was
applied to all three. When CSD indicators did appear, they only did so
in the juvenile population variables. This represents another
under-explored area - selecting appropriate variables for early-warning
detection in multivariate systems. Even where CSD is present, it may not
be expressed in all system components.

@@

CSD precedes several types of these non-catastrophic bifurcations. In
the subcritical form of a Hopf bifurcation, a system transitions from a
stable equilibrium to a stable cycle. As a control parameter approaches
the critical threshold, the system's dominant eigenvalue approaches zero
and thus exhibits CSD (Chisholm and Filotas 2009; Kéfi et al. 2012).
However, the mean value of the equilibrium does not change dramatically,
and the transition from stable equilibrium to cycles is gradual as the
cycle sizes grow from zero at the threshold value. To appreciate how
this bifurcation is gradual rather than catastrophic, note that in the
presence of stochasticity, the system behavior observed on either side
of the threshold may be indistinguishable: on one side stochasticity
bounces the system around a stable node, while on the other it bounces
the system around a very small limit cycle in the same region of state
space. Contrast this to a critical transition in which any stochastic
fluctuation across the threshold could lead to a qualitatively different
state.

The system's eigenvalue also passes through zero in the case of the
transcritical bifurcation. The transcritical is a degenerate case of the
saddle-node, and occurs in many of the same systems. However, when a
system passes through a transcritical bifurcation, the stable
equilibrium transitions smoothly from positive to zero, or the reverse.
In population systems, this corresponds to a transition from an
equilibrium of a very small population size to extinction - an important
but non-catastrophic, and probably directly observable, event. CSD is
observed prior to the transcritical bifurcations (Chisholm and Filotas
2009; Kéfi et al. 2012).

@@

\begin{figure}[htbp]
\centering
\includegraphics{kefi-fig.pdf}
\caption{A system where critical slowing down is observed without a
critical threshold, from Kéfi et al. (2012). In this model, prey have
logistic growth and are subject to predation with a Type III functional
response, but there is no bifurcation. Instead, average prey population
exhibits a smooth response to increased predation (grazing). Left: The
population level as a function of predation rate. Middle: Variance of
the prey population level. Right: Lag-1 Autocorrelation in prey
population dynamics as grazing rate increases. Note that both indicators
increase despite the lack of a bifurcation.}
\end{figure}

@@

Some rapid regime shifts are not due to bifurcations at all. A large
external forcing (as illustrated in Figure 4) may change the behavior of
a system without any warning. This mechanism is commonly recognized,
(Barnosky et al. 2012; Scheffer et al. 2001; Scheffer et al. 2009;
Scheffer et al. 2012), but others are possible. An internal stochastic
event may switch a system between dynamic regimes, or a change in system
behavior may be the manifestation of a long-term transient. In none of
these cases would CSD be expected to precede such changes.

\begin{figure}[htbp]
\centering
\includegraphics{Bel2012example.pdf}
\caption{Difference between different types of perturbations. On the
horizontal axis is the bifurcation parameter, representing the state of
the environment (e.g.~annual mean temperature) whose slow change could
lead to a sudden shift. A direct disturbance to the system state
(e.g.~population size, vertical axis) could also cause a transition if
it is large enough to cross the stability threshold (dashed line). Such
a perturbation can come from exogenous factors such as anthropogenic
pressures or occur by chance from intrinsic stochasticity. These
distinct mechanisms of disturbance and environmental change are coupled
-- as the environment deteriorates, moving the system right on the
diagram, the probability that a disturbance crosses the threshold
increases. From Bel, Hagberg, and Meron (2012).}
\end{figure}

Large, rapid changes in external conditions will result in rapid changes
in ecological system dynamics. For instance, rapid changes in North
American vegetation at the start of the Bølling-Allerød and end of the
Younger Dryas period are thought to be responses to similarly large,
rapid changes in climate (Williams, Blois, and Shuman 2011). Doney and
Sailley (2013) interpret a recent analysis by Di Lorenzo and Ohman
(2013) as demonstrating that what were previously thought of as regime
shifts in krill dynamics in the Pacific ocean (Hare and Mantua 2000)
could actually be explained by a close coupling to the external forcing
of El Nino environmental dynamics through the Pacific Decadal
Oscillation (PDO). Schooler et al. (2011) found that lakes with the
invasive plant \emph{Salvaniai molesta} andherbivorous weevils
alternated between low- and high-\emph{Salvnia} states driven by
disturbances from regular external flooding events. These examples
highlight cases that involve critical transitions between regimes under
circumstances that do not permit the discovery of early warning signals,
as CSD is not anticipated under these mechanisms.

>>>>

\documentclass{article} 
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.

\title{Early warning signals: The charted and uncharted territories}
\author{Carl Boettiger, Noam Ross, Alan Hastings}
\begin{document}

@@

The realization that complex systems such as ecological communities can collapse or shift regimes suddenly and without rapid external forcing poses a serious challenge to our understanding and management of the natural world.  The potential to identify early warning signals that would allow researchers and managers to predict such events before they happen has therefore been an invaluable discovery that offers a way forward in spite of such seemingly unpredictable behavior.  Research into early warning signals has demonstrated that it is possible to define and detect such early warning signals in advance of a transition in certain contexts.  Here we describe the pattern emerging as research continues to explore just how far we can generalize these results.  A core of examples emerges that shares three properties: the phenomenon of rapid regime shifts,  a pattern of 'critical slowing down' that can be used to detect the approaching shift, and a mechanism of bifurcation driving the sudden change.  As research has expanded beyond these core examples, it is becoming clear that not all systems that show regime shifts exhibit critical slowing down, or vice versa.  Even when systems exhibit critical slowing down, statistical detection is a challenge.  We review the literature that explores these edge cases and highlight the need for (a) new early warning behaviors that can be used in cases where rapid shifts do not exhibit critical slowing down, (b) the development of methods to identify which behavior might be an appropriate signal when encountering a novel system; bearing in mind that a positive indication for some systems is a negative indication in others, and (c) statistical methods that can distinguish between signatures of early warning behaviors and noise.
\\
Key words: early warning signals, regime shifts, bifurcation, critical slowing down 
\\
\end{abstract}
%\footnote[*]{authors contributed equally}
%\footnote[a]{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA}
%\footnote[b]{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA}

@@

Many natural systems exhibit regime shifts - rapid changes in the state
and conditions of system behavior. Such shifts include lake
eutrophication (Carpenter et al. 1999), algal overgrowth of coral
systems (Mumby et al. 2007), fishery collapse (Jackson et al. 2001),
desertification of grasslands (Kéfi et al. 2007), and rapid changes in
climate (Dakos et al. 2008, Lenton et al. 2009). Such dramatic shifts
have the potential to impact ecosystem health and human well-being.
Thus, it is important to develop strategies for adaptation, mitigation,
and avoidance of such shifts.

The idea that complex systems such as ecosystems could change suddenly
and without warning goes back to the 1960s (Lewontin 1969, Holling 1973,
May 1977). Such early work revealed that even simple models with the
appropriate nonlinearities were capable of unpredictable behavior. The
only way to predict the transition was to have the right model -- and
that meant having already had the chance to observe the transition (even
then, this remains a tough problem). One cogent early example (Ludwig et
al. 1978) demonstrated how knowledge of the forms and time scales of
interactions among insects, birds, and trees could lead to a qualitative
model that essentially predicted the possibility of regime shifts.

@@

Foundational research in EWS identified certain patterns that may
forecast a sudden transition in a wide variety of systems (Scheffer et
al. 2009). Most extensively researched is the phenomenon of critical
slowing down (CSD), which is manifested as a pattern of increasing
variance or autocorrelation of a system. Subsequent work has begun to
identify a growing library of cases in which these indicators are not
present before a transition (Schreiber and Rittenhouse 2004, Schreiber
and Rudolf 2008, Hastings and Wysham 2010, Bel et al. 2012), or are
observed in the absence of any transition (Kéfi et al. 2012). These
examples are distinct from the more well-known case of statistical error
-- such as a signal that is present but too weak to detect due to
insufficient available data (see Dakos et al. (2008); Scheffer et al.
(2009) and Perretti and Munch (2012)). Instead, such work moves into new
territory where different underlying mechanisms have lead to starkly
different patterns. Determining which underlying mechanisms are present
is a substantial empirical and theoretical challenge. When does critical
slowing down correspond to the assumptions made?

Here we review a variety of mechanisms that may lead to rapid (or
``catastrophic'') regime shifts in ecological systems, as well as
mechanisms that generate early warning signals. We focus on CSD and its
manifestations as they are the most commonly studied warning signals. We
illustrate that not all rapid shifts exhibit CSD, and not all
observations of CSD involve rapid shifts. Thus the issue of determining
EWS is really two-fold: first, to identify classes of systems where the
warning signal is expected and conversely systems that may undergo
shifts without such signals, and second, to determine appropriate
statistical tools to detect the warning signal. In this paper we review
both aspects of the overall question.

\begin{longtable}[c]{@{}ll@{}}
\hline\noalign{\medskip}
\begin{minipage}[t]{0.42\columnwidth}\raggedright
\textbf{Critical slowing down (CSD)}
\end{minipage} & \begin{minipage}[t]{0.55\columnwidth}\raggedright
A system's slowing response to perturbations as it's dominant eigenvalue
approaches zero, often expressed in greater variance, autocorrelation,
and return time. CSD is one possible EWS.
\end{minipage}
\\\noalign{\medskip}
\begin{minipage}[t]{0.42\columnwidth}\raggedright
\textbf{Early warning signals (EWS)}
\end{minipage} & \begin{minipage}[t]{0.55\columnwidth}\raggedright
A general term for dynamic patterns in system behavior that precede
regime shifts. Though CSD phenomena are among the best studied EWS, some
shifts will require alternative signals; Figure 1.
\end{minipage}
\\\noalign{\medskip}
\hline
\end{longtable}

@@

CSD has been studied extensively in theoretical (Wissel 1984, Gandhi et
al. 1998, Carpenter and Brock 2006, Hastings and Wysham 2010, Dakos et
al. 2011a, Lade and Gross 2012, Boettiger and Hastings 2012a) and
empirical contexts (Drake and Griffen 2010, Carpenter 2011a, Veraart et
al. 2012, Dai et al. 2012, Wang et al. 2012) as a potential EWS for
regime shifts. CSD occurs as a system's dominant eigenvalue approaches
zero due to a changing (possibly deteriorating) environment. As the
eigenvalue approaches zero, the system's response to small perturbations
slows. This change in dynamic properties of a system can be expressed in
greater variance, autocorrelation, and return time of observed state
variables.

@@

\begin{figure}[htbp]
\centering
\includegraphics{ews-venn.eps}
\caption{Venn diagram representing the intersecting domains of rapid
regime shifts, bifurcations, and critical slowing down. Labels in italic
are example phenomena that occur in each domain. Roman numerals indicate
example literature (right) exploring each domain, and also refer to
sections below describing those domains. The center domain (I) where all
three phenomena intersect, is the most extensively researched domain of
the EWS field. Literature outside this charted region does not yet
provide the needed EWS, but hints where existing signals based on CSD
may be insufficient or misleading.}
\end{figure}

@@

Much of the (most visible) recent research in EWS has focused on the
center of the diagram, where all three concepts intersect. The warning
signal patterns postulated, such as increasing variance and coefficient
of variation, (Carpenter and Brock 2006), increasing autocorrelation
(Dakos et al. 2008), increasing skewness (Guttal and Jayaprakash 2008a)
can all be directly derived from the changing eigenvalue in a saddle
node (also called fold) bifurcation. Consequently, experimental
evaluations of warning signals have largely focused on this situation as
well. CSD has frequently been studied in the context of models
exhibiting saddle-node bifurcations.

Dai et al. (2012) studied yeast cell growth in a microcosm and
demonstrated that an Allee effect created a saddle-node bifurcation in
the system. When the cell density was reduced to levels near the
bifurcation point, a decrease in recovery time (increase in variance and
autocorrelation over time) was observed. Veraart et al. (2012) studied a
system of cyanobacteria where models suggest a saddle-node bifurcation
driven by light inhibition. They also found increases in autocorrelation
and decreased recovery rates as the system approached the bifurcation.
These important experiments are among the best demonstrations that
saddle-node bifurcation dynamics really occur in natural systems, and
can be accompanied by reliable detection of EWS, at least when
sufficient data sampling, replicates, and controls are available.

Carpenter et al. (2011) provide a larger-scale example in which a lake
ecosystem is manipulated towards a sudden transition through the
introduction of a predator, while a neighboring experimental lake
provides a control. In this and similar lake systems, bifurcation is
thought to be driven in part by trophic interactions where adult fish
prey on the competitors of their juveniles (Carpenter and Kitchell 1996,
Walters2001; Carpenter et al. 2008) which leads to a saddle-node
bifurcation. While the underlying dynamics of a whole lake ecosystem are
less tractable than the laboratory controlled chemoststats of
microorganisms, the system is understood well enough to anticipate that
a sudden transition can be induced under the intended manipulation. Like
the laboratory examples, this helps eliminate the options outside the
circle ``bifurcations,'' in Figure 1. The observed warning signals then
place it in the center of the diagram.

@@

These are some of the most problematic cases. They represent disruptive
but potentially avoidable events, but would not be detected by using CSD
as an EWS. These cases include bifurcations in continuous time
(Schreiber and Rudolf 2008) and discrete time (Schreiber 2003),
explicitly spatial (Bel et al. 2012) and non-spatial, chaotic (Schreiber
2003, Hastings and Wysham 2010) and non-chaotic (Schreiber and Rudolf
2008, Hastings and Wysham 2010, Bel et al. 2012) examples. Before
warning signals can be reliably applied to novel systems, research must
provide a way to discern if the dynamics correspond to the better
understood warning signals of the saddle-node case or the more complex
patterns such as the examples discussed here.

One class of bifurcations in which we would not expect to see CSD prior
to regime shift are sometimes known as \emph{crises}. Crises are sudden
changes in the dynamics of chaotic attractors that occur in response to
small changes in parameters (Grebogi et al. 1983). Chaotic attractors
are features of many ecological models (Hastings et al. 1993), and
chaotic behavior has been shown in some ecological systems (Costantino
et al. 1997).

@@

\begin{figure}[htbp]
\centering
\includegraphics{schreiber-fig.eps}
\caption{A system where variance decreases prior to a population
collapse; adapted from Schreiber (2003). In this model, prey species
with high growth rates exhibit chaotic dynamics under predation, but
populations collapse when predation increases beyond a threshold value.
Left: The population level as a function of predation rate. Mean
dynamics shown as black line, realizations with varying initial
conditions shown as grey dots; see Schreiber (2003). Middle: Variance of
the prey population level. Note that it \emph{decreases} as predation
rate approaches the threshold. Right: Lag-1 Autocorrelation in prey
population dynamics increases as the threshold is approached}
\end{figure}

@@

Another non-chaotic example is found in some spatially extended systems
that exhibit a type of bifurcation not accompanied by CSD. In this class
of models, individual locations are subject to saddle node-type regime
shifts and influence adjacent locations via short-range facilitation and
long-range competition. Such models are used represent transitions
between vegetation types in response to changing water availability, and
reproduce naturally occurring vegetation patterns (Rietkerk and van de
Koppel 2008). In such systems, a regime shift in one location can
propagate spatially and transition the whole system from one regime to
another. Such a transition occurs if the control parameter (e.g.,
rainfall), exceeds the \emph{Maxwell point} - the value at which a local
disturbance propagates outwards (Bel et al. 2012). The Maxwell point may
be far from the level at which an individual location would undergo a
saddle-node bifurcation, and thus the system's global dynamics would not
exhibit CSD prior to such a transition. This case illustrates the
importance of distinguishing between \emph{local} and \emph{global}
system dynamics and identifying the appropriate scale of observation.

Finally, Boerlijst et al. (2013) found that indicators of CSD do not
appear prior to saddle-node bifurcations when perturbations are not in
the direction of a system's dominant eigenvalue, and even then may only
appear in one variable of the system. In their example case, increased
variance and autocorrelation only occurred when noise was applied to the
juvenile population of a model with juveniles, adults, and predators,
and it did not appear when identical noise was applied to all three.
When CSD indicators did appear, they only did so in the juvenile
population variables. This represents another under-explored area -
selecting appropriate variables for early-warning detection in
multivariate systems. Even where CSD is present, it may not be expressed
in all system components.

@@

CSD precedes several types of these non-catastrophic bifurcations. In
the subcritical form of a Hopf bifurcation, a system transitions from a
stable equilibrium to a stable cycle. As a control parameter approaches
the critical threshold, the system's dominant eigenvalue approaches zero
and thus exhibits CSD (Chisholm and Filotas 2009, Kéfi et al. 2012).
However, the mean value of the equilibrium does not change dramatically,
and the transition from stable equilibrium to cycles is gradual as the
cycle sizes grow from zero at the threshold value. To appreciate how
this bifurcation is gradual rather than catastrophic, note that in the
presence of stochasticity, the system behavior observed on either side
of the threshold may be indistinguishable: on one side stochasticity
bounces the system around a stable node, while on the other it bounces
the system around a very small limit cycle in the same region of state
space. Contrast this to a critical transition in which any stochastic
fluctuation across the threshold could lead to a qualitatively different
state.

The system's eigenvalue also passes through zero in the case of the
transcritical bifurcation. The transcritical is a degenerate case of the
saddle-node, and occurs in many of the same systems. However, when a
system passes through a transcritical bifurcation, the stable
equilibrium transitions smoothly from positive to zero, or the reverse.
In population systems, this corresponds to a transition from an
equilibrium of a very small population size to extinction - an important
but non-catastrophic, and probably directly observable, event. CSD is
observed prior to the transcritical bifurcations (Chisholm and Filotas
2009, Kéfi et al. 2012).

@@

\begin{figure}[htbp]
\centering
\includegraphics{kefi-fig.eps}
\caption{A system where critical slowing down is observed without a
critical threshold, from Kéfi et al. (2012). In this model, prey have
logistic growth and are subject to predation with a Type III functional
response, but there is no bifurcation. Instead, average prey population
exhibits a smooth response to increased predation (grazing). Left: The
population level as a function of predation rate. Middle: Variance of
the prey population level. Right: Lag-1 Autocorrelation in prey
population dynamics as grazing rate increases. Note that both indicators
increase despite the lack of a bifurcation.}
\end{figure}

@@

Some rapid regime shifts are not due to bifurcations at all. A large
external forcing (as illustrated in Figure 4) may change the behavior of
a system without any warning. This mechanism is commonly recognized,
(Scheffer et al. 2001, 2009, Barnosky et al. 2012, Scheffer et al.
2012), but others are possible. An internal stochastic event may switch
a system between dynamic regimes, or a change in system behavior may be
the manifestation of a long-term transient. In none of these cases would
CSD be expected to precede such changes.

\begin{figure}[htbp]
\centering
\includegraphics{Bel2012example.eps}
\caption{Difference between different types of perturbations. On the
horizontal axis is the bifurcation parameter, representing the state of
the environment (e.g.~annual mean temperature) whose slow change could
lead to a sudden shift. A direct disturbance to the system state
(e.g.~population size, vertical axis) could also cause a transition if
it is large enough to cross the stability threshold (dashed line). Such
a perturbation can come from exogenous factors such as anthropogenic
pressures or occur by chance from intrinsic stochasticity. These
distinct mechanisms of disturbance and environmental change are coupled
-- as the environment deteriorates, moving the system right on the
diagram, the probability that a disturbance crosses the threshold
increases. From Bel et al. (2012).}
\end{figure}

Large, rapid changes in external conditions will result in rapid changes
in ecological system dynamics. For instance, rapid changes in North
American vegetation at the start of the Bølling-Allerød and end of the
Younger Dryas period are thought to be responses to similarly large,
rapid changes in climate (Williams et al. 2011). Doney and Sailley
(2013) interpret a recent analysis by Di Lorenzo and Ohman (2013) as
demonstrating that what were previously thought of as regime shifts in
krill dynamics in the Pacific ocean (Hare and Mantua 2000) could
actually be explained by a close coupling to the external forcing of El
Nino environmental dynamics through the Pacific Decadal Oscillation
(PDO). Schooler et al. (2011) found that lakes with the invasive plant
\emph{Salvaniai molesta} andherbivorous weevils alternated between low-
and high-\emph{Salvnia} states driven by disturbances from regular
external flooding events. These examples highlight cases that involve
critical transitions between regimes under circumstances that do not
permit the discovery of early warning signals, as CSD is not anticipated
under these mechanisms.
